Prediction of Protein-Protein Interactions on the Human and Rice Interactome

### 2020-09-24 nicolas/PPIPL3/ReportV3.tex
Report - Version 3

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

@@

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactomes}
\author{Nicolás Antonio López Rozo}
\maketitle
\begin{abstract}

Network analysis for  predict unmapped protein-protein interactions (PPI) suggest that the higher the number of paths of
length~3 (L3) between two proteins, the more likely they are to interact. This paper extends previous work based on the L3 principle by taking into account the representation learning of node features of the PPI network. In particular, we train an XGBoost model using L3 and handcrafted features, as well as embeddings from \texttt{Node2Vec}. Our main result shows that while L3 is an important feature for predicting links, best performance is achieved when combined with edge embeddings. The proposed approached is evaluated for the human and rice interactomes.
\end{abstract}
\begin{comment}
need include a feature that compares the dimension of the embedding
is useful for prediction with L3 to show that they independent measures
- we do not want to be learning L3 through node2vec
\end{comment}

@@

Proteins are the key actors of biological processes inside cells. As they carry out a variety of tasks, cells do not work as single agents but
rather as a part of dynamic networks of protein-protein interaction (PPI)~\cite{Lin2017}. PPI networks play a key role in a variety of interdependent mechanisms, including signal transduction, homeostasis control, stress responses, plant defense and organ formation. At the molecular level, PPIs play an essential role in many physiological and developmental processes, including protein phosphorylation, transcriptional co-factor recruitment, transporter activation~\cite{Zhang_2010_PPI}.

@@

A most common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid} technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on (i) the expression of a specific reporter gene that is activated by the binding of a DNA-binding Domain (DB); and (ii) an activation domain (AD) of a transcription factor that binds to an upstream activation sequence (UAS). For the Y2H technique, a protein is fused to the DB domain (known as \emph{bait}) and another one to the AD (known as \emph{prey}). If the proteins do not interact, then the reporter gene
is not expressed. Otherwise, the reporter gene expression is activated by the activation domain.

\begin{comment}
NL: I will make this figure myself
JF: Good idea
\end{comment}

@@

Based on the outcome of numerous Y2H experiments PPI networks can be constructed, in which known interactions for each protein are represented. Several algorithms are proposed over these networks in order to predict yet unknown interactions. This paper evaluates predictions based on three measures: Common Neighbors (\textbf{CN}), based on the count of paths of length 2; the raw count of paths of length 3
(\textbf{A3}); and the degree-normalized count of paths of length 3(\textbf{L3}).

\begin{comment}
JF: Yet you do more than evaluating this three measures so this has to be updated. 
JF: Not sure we should use L3 above since it is being redefined as a normalized measure here!
\end{comment}

The focus of this paper is to evaluate different methods for predicting PPIs using the existing knowledge of the network of interactions, which is represented as an undirected graph. The traditional way is of approaching the problem is usually based on social networks analysis, more specifically on the triadic closure principle (TCP), that states that the more common shared friends that two people have, the more likely that they know each other. As shown by previous studies, the mentioned approach fails because it does not consider the structural and chemical properties
of the proteins \cite{Kovacs2019}.

\begin{comment}
JF: Not sure this is the place to introduce TCP. You are already stating what the paper will do. Sees like TCP, should be introduced before the L3 measure is introduces, as part of the history of past approaches. Also TCP is not referred to later on...
\end{comment}

@@

%% read throughout the document but edited up here. I prefer to read the results before editing the materials and methods section. 
\section{Materials and Methods}

@@

Information for the human interactome was used as-is, which corresponds
to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
$9433$ and $23322$ interactions.

@@

For each network of protein interactions, the following procedure
was performed 10 times in order to address the stochastic nature of
the process and have a consensus:
\begin{itemize}
\item A percentage of interactions is removed at random from the network
(20\%).
\item The same amount of removed interactions are then predicted using the
main methods for prediction mentioned by Kovacs et al (2019): Common
Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
and the Length-3 degree-normalized score (\textbf{L3}).
\item A test dataset is created as follows: all removed edges are included
(as observed positives for the ML algorithm) and from the predicted
edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
the previous classification (observed negatives), a random subset
is chosen such that the dataset is balanced, that is, the amount of
observed positive labels is equal to the observed negative labels.
\item Once the dataset is ready, it is randomly partitioned: 80\% is used
for \texttt{XGBoost} model training and 20\% is used for validation.
It is important to have in mind that balanced distribution of the
positive and negative labels in the datasets was satisfied.
\end{itemize}

@@

For the rice interactome, the different model-features combinations
were trained and validated. After executing the mentioned routines,
the results are shown in Figure \ref{F1}. First, one should have
a baseline of comparison, which in this case corresponds to \texttt{Node2Vec}
without any additional feature included. The plot below shows those
results, and one can see that its mean performance using the AUC metric
is 0.52, and that the results among the 10 repetitions are consistent.
This results mean that the model using only the default features perform
barely as good as a random choice of the labels.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F1}Summary ROC curves for \texttt{Node2Vec} model alone}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.95\columnwidth]{Only_ML/ROConlyML_SUMMARY}
\end{figure}

\section{(PENDING FROM HERE ON)}

% cannot compile doc given the way images are being imported

%\begin{figure}[h]
%\caption{Methods Comparison for \emph{HI-II-14}}
%
%\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
%\end{figure}
%
%\begin{figure}[h]
%\caption{Methods Comparison for \emph{HI-TESTED}}
%
%\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
%\end{figure}

@@

When analyzing the robustness of each of the networks, the following
values for the Weighted Spectral Distribution were found. For a robustness
reference, the Erdos-Renyi model was used to generate a random network
with the same number of nodes and edges and on those random networks,
the WSD was measured.
\begin{center}
\begin{table}[H]
\caption{Validation of Weighted Spectral Distribution}

\centering{}%
\begin{tabular}{|c|c|c|}
\hline 
Network & Network WSD & Erdos-Renyi WSD\tabularnewline
\hline 
\hline 
HI-II-14 & 393.4939 & 198.7706\tabularnewline
\hline 
HI-TESTED & 423.3902 & 276.1065\tabularnewline
\hline 
HI-III (VALID.) & 373.9369 & 153.5329\tabularnewline
\hline 
\end{tabular}
\end{table}
\par\end{center}

The table shows that the three networks used in this report are robust,
because the are significantly more robust than a network with the
same density generated randomly.

@@

\begin{comment}
please update the conclusions - WSD does not seem relevant
\end{comment}

For the analyzed networks, it is also important to highlight their
robustness when compared to a random network with the same density:
if weighted spectral distribution (WSD) is used, then the used networks
are on average twice as robust as their random network versions.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

@@

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
\author{Nicolás Antonio López Rozo}
\maketitle
\begin{abstract}

Network analyses for predicting unmapped protein-protein interactions
(PPI) suggest that the higher the number of paths of length\textasciitilde 3
(L3) between two proteins, the more likely they are to interact. This
paper extends previous work based on the L3 principle by taking into
account the representation learning of node features of the PPI network.
In particular, we train an \texttt{XGBoost} model using L3 and handcrafted
features, as well as embeddings from \texttt{Node2Vec}. Our main result
shows that while L3 is an important feature for predicting links,
best performance is achieved when combined with edge embeddings. The
proposed approach is evaluated for the human and rice interactomes.
\end{abstract}
\begin{comment}
need include a feature that compares the dimension of the embedding
is useful for prediction with L3 to show that they independent measures
- we do not want to be learning L3 through node2vec
\end{comment}

@@

Proteins are the key actors of biological processes inside cells.
As they carry out a variety of tasks, proteins do not work as single
agents but rather as a part of dynamic networks of protein-protein
interactions (PPI) \cite{Lin2017}. PPI networks play a key role in
a variety of interdependent mechanisms, including signal transduction,
homeostasis control, stress responses, plant defense and organ formation.
At the molecular level, PPIs play an essential role in many physiological
and developmental processes, including protein phosphorylation, transcriptional
co-factor recruitment and transporter activation \cite{Zhang_2010_PPI}.

@@

A common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid}
technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on the expression of a specific reporter gene that
activates by the binding of a DNA-binding Domain (DB) and an Activation
Domain (AD) of a Transcription Factor that binds to an Upstream Activation
Sequence (UAS). For the Y2H technique, a protein is fused to the DB
domain (known as \emph{bait}) and another one to the AD (known as
\emph{prey}). If the proteins do not interact, then the reporter gene
is not expressed. Otherwise, the reporter gene expression is activated
by the activation domain.

\begin{comment}
NL: I will make this figure myself

JF: Good idea
\end{comment}

@@

Based on the outcome of numerous Y2H experiments PPI networks can
be constructed, in which known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
yet unknown interactions. This paper evaluates predictions based on
three measures: Common Neighbors (\textbf{CN}), based on the count
of paths of length 2; the raw count of paths of length 3 (\textbf{A3});
and the degree-normalized count of paths of length 3 (\textbf{L3}).
Furthermore

\begin{comment}
JF: Yet you do more than evaluating this three measures so this has
to be updated.

JF: Not sure we should use L3 above since it is being redefined as
a normalized measure here!
\end{comment}

The focus of this paper is to evaluate different methods for predicting
PPIs using the existing knowledge of the network of interactions,
which is represented as an undirected graph. The traditional way of
approaching the problem is usually based on social networks analysis,
more specifically on the triadic closure principle (TCP), that states
that the more common shared friends that two people have, the more
likely that they know each other. As shown by previous studies, the
mentioned approach fails because it does not consider the structural
and chemical properties of the proteins \cite{Kovacs2019}.

@@

\begin{comment}
JF: read throughout the document but edited up here. I prefer to read
the results before editing the materials and methods section. 
\end{comment}

\section{Materials and Methods}

@@

Information for the human interactome was used as-is, which corresponds
to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
$9433$ and $23322$ interactions, respectively.

@@

For each network of protein interactions, the following procedure
was performed 10 times in order to address the stochastic nature of
the process and have a consensus:
\begin{itemize}
\item A percentage of interactions is removed at random from the network
(20\%).
\item The same amount of removed interactions are then predicted using the
main methods for prediction mentioned by Kovacs et al (2019): Common
Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
and the Length-3 degree-normalized score (\textbf{L3}).
\item A test dataset is created as follows: all removed edges are included
(as observed positives for the ML algorithm) and from the predicted
edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
the previous classification (observed negatives), a random subset
is chosen such that the dataset is balanced, that is, the amount of
observed positive labels is equal to the observed negative labels.
\item Once the dataset is ready, it is randomly partitioned: 80\% is used
for \texttt{XGBoost} model training and 20\% is used for validation.
It is important to have in mind that balanced distribution of the
positive and negative labels in the datasets was satisfied.
\end{itemize}
It is worth mentioning that some exploration experiments were carried
out with interactions removal percentages of 2, 5, 10 and 20\%, as
well as train-test partitions of 15-85, 80-20 and 75-25, and the explained
parameter set was selected because it either represented a marginal
gain (results not shown) or because it was a common parameter selection
in related literature.

@@

For the rice interactome, the different model-features combinations
were trained and validated. After executing the mentioned routines,
the results are shown in Figure \ref{F1}. First, one should have
a baseline of comparison, which in this case corresponds to \texttt{Node2Vec}
without any additional feature included. The plot below shows those
results, and one can see that its mean performance using the AUC metric
is $0.52$, and that the results among the 10 repetitions are consistent.
This results mean that the model using only the default features perform
barely as good as a random choice of the labels. This result can also
be assessed when looking at the confusion matrix, where a precision
of $0.5168$ and a recall of $0.4589$ can be derived.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F1}Summary ROC curves for \texttt{Node2Vec} model alone}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_ML/ROConlyML_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_ML/CMonlyML_SUMMARY}
\end{figure}

Next experiment carried out was to add each prediction method score
as a prediction feature, that is, use the score as the eleventh input
feature for the ML algorithm. Results for the Common Neighbors method
(\textbf{CN}), which uses paths of length 2, are shown in Figure \ref{F2}.
It can be observed again that all 10 experiments have small variability
among them and perform significantly better than the baseline, with
an area under the ROC curve of $0.90$, with small standard deviation.
When looking at the confusion matrix for this model, turns out that
most of the guesses are true positives and true negatives, resulting
in a precision of $0.9285$ and a recall of $0.7420$. Same analyses
are done with the count of paths of length 3 (\textbf{A3}) and with
the degree-normalized length-3 score (\textbf{L3}) and results are
presented in figures \ref{F3} and \ref{F4}, respectively, resulting
also in area under the curve of 0.9 for both cases.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F2}Results for \texttt{Node2Vec} model with CN feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA2_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F3}Results fir \texttt{Node2Vec} model with A3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA3_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F4}Results for \texttt{Node2Vec} model with L3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CML3_SUMMARY}
\end{figure}

The precision when using the count of paths of length 3 (A3) is $0.8587$
while the recall is $0.7685$. It is interesting to observe that precision
decreased $7.5\%$ when compared to the previous case, but recall
increased $3.6\%$. A similar situation is obtained for the model
with the normalized score (L3), whose precision decreased to $0.8936$
($-3.8\%$) and recall increased to $0.7610$ ($+2.6\%$). It can
be seen also that the ROC curve rapidly achieves higher values of
true positive rate in the A2-featured model when compared to the other
models.

Finally, an relevant evaluation on the models that include a handcrafted
feature is necessary: How important is the appended feature for the
model result? The answer can be observed by the feature importance
plots for CN, A3 and L3 in Figure \ref{F5-importance}. Each plot
resembles the number of times that each feature creates a bifurcation
in the underlying decision trees that \texttt{XGBoost} uses. The more
bifurcations, the higher the importance a feature has on the model
itself. All models have the appended feature as the most important
by a large margin.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F5-importance}Importance for \texttt{Node2Vec} models: with
L3 feature}
\par\end{centering}
\noindent \begin{raggedleft}
\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot 1}
\par\end{raggedleft}
\noindent \begin{raggedleft}
\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot 1}
\par\end{raggedleft}
\noindent \raggedleft{}\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot 1}
\end{figure}

\subsection{Human Interactome}

(PENDING FROM HERE)

\begin{figure}[h]
\caption{Methods Comparison for \emph{HI-II-14}}

\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-ii-14\lyxdot txt}
\end{figure}

\begin{figure}[h]
\caption{Methods Comparison for \emph{HI-TESTED}}

\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-tested\lyxdot txt}
\end{figure}

@@

@@

\begin{comment}
JF: please update the conclusions - WSD does not seem relevant

NL: Done
\end{comment}
